A Reduced Order Model for Permeability Fields Using Singular Value Decomposition, SVD, by Interpolating in the Principal Component Spaces \red{(Maybe you should consider a shorter and punchier title ...)}

### 2018-03-26 Papers/AppliedMaths2018/Stochastic_Upscaling_UK_AMM_2018.tex
Updates

\newcommand{\JGnote}[1]{\fbox{\parbox{\textwidth}{ \color{red} JG Note $\Rightarrow$ #1}}}
\newcommand{\BLnote}[1]{\fbox{\parbox{\textwidth}{ \color{green} BL Note $\Rightarrow$ #1}}}
\newcommand{\red}{\textcolor{red}}
\newcommand{\blue}{\textcolor{blue}}
\newcommand{\green}{\textcolor{green}}
\newcommand{\frc}{\displaystyle\frac}
\newcommand{\PN}[2][error]{P$_{#1}$DG-P$_{#2}$}
\newcommand{\PNDG}[2][error]{P$_{#1}$DG-P$_{#2}$DG}
\newcommand{\eg}{{\it e.g., }} 
\newcommand{\ie}{{\it i.e., }}

@@

In Section 2, the basic mathematical explanation of SVD and its associated properties are presented. Then, the concept of spaces is explained from the engineer's perspective and also from the mathematician's perspective. The objective of this explanation is to assist both parties in visualizing the relevant spaces and performing the necessary transformation within those spaces. Finally, the section describes model order space reduction for the permeability field within the principal component space. Section 3, starts by describing the high resolution base case on which the four upscaling techniques are modelled. The the simulation run which is similar for all the cases/models is described. A brief summary of the pre-processing step for each of the models is also presented. Section 4 provides the results and some discussion on the results of the simulations. It als gives further intepretation to the result obtained from the SVD upscaling technique. Finally, section 5 provides the conclusion.

\section{Singular Value Decomposition, SVD, and its application for reducing the order of a permeability field}\label{section:svd}
\subsection{A Brief Introduction to SVD}\label{subsection:svd_brief}

Singular Value Decomposition, SVD, \JGnote{ioioioiok} exists within the linear algebra branch of mathematics as one of the most exciting means of factorizing a matrix, $\mathbf{A}$, into three parts $\mathbf{U}$, $\Sigma$ and $\mathbf{V}$. Where $\mathbf{U}$ and $\mathbf{V}$ are orthonormal, and $\Sigma$ is a diagonal matrix. If $\mathbf{A}$, is a square matrix then $\mathbf{U}$, $\Sigma$ and $\mathbf{V}$ are square matrix and $\mathbf{U}^{\intercal}\mathbf{U} = \mathbf{V}^{\intercal}\mathbf{V} = \mathbf{I}$. But if $\mathbf{A}$ is an $m * n$ matrix, $\mathbf{U}$ and $\mathbf{V}$ still remain as square matrices, albeit with different matricial sizes. $\mathbf{U}$ is an $m * m$ while $\mathbf{V}$ is an $n * n$. Another interesting property of $\mathbf{U}$ and $\mathbf{V}$ since they are square (irrespective of whether A is square or rectangular) is that $\mathbf{U}^{-1} = \mathbf{U}^{\intercal}$ and $\mathbf{V}^{-1} = \mathbf{V}^{\intercal}$. When $\mathbf{A}$ is an $m * n$ matrix, $\Sigma$ is a diagonal $m x n$ matrix. The diagonal entries of $\Sigma$ (in both the case of square and rectangular matrices) are referred to as the singular values of $\mathbf{A}$, and are usually denoted by sigma ($\sigma_{1}, \sigma_{2},..., \sigma_{r}$). For a rectangular matrix, the singular values fill the first $r$ places on the main diagonal of $\Sigma$, where $r$ is the rank of A. Additionally, the $i$th singular value, $\sigma_{i}$ is ordered such that $\sigma_{1} \geq \sigma_{2} \geq ...\geq \sigma_{r}$. A huge benefit of the SVD is the orthogonality of the columns in $\mathbf{U}$ and $\mathbf{V}$. The columns of $\mathbf{U}$ are known as the left singular vectors while the columns in $\mathbf{V}$ are known as the right singular vectors. A simple consideration of the benefit of this, is the ease with which it allows the inverse of square matrix $\mathbf{A}$ to be calculated. This example is presented thus:

\begin{equation}
 \mathbf{A}x = b
\end{equation}
\begin{equation}
 \mathbf{U} \Sigma \mathbf{V}^{\intercal} x = b
\end{equation}
\begin{equation}
 x = \mathbf{V} \Sigma^{-1} \mathbf{U}^{\intercal} b
\end{equation}

\subsection{Principal Component Spaces and Model Order Reduction for the permeability field}\label{subsection:pcspaces}

\subsubsection{Visualizing the permeability space}\label{subsubsection:visualization_permspace}
In real-life ({\it i.e.}, a 3D geometry), permeability is a tensor quantity, but for the purpose of the following explanation (and the remaining part of this paper), a 2D geometry is assumed where permeability is a scalar quantity. Furthermore, the simulator prescribes the permeability field using $P0DG$ element type, which means that permeability values are assigned in the middle node of the triangular unstructured mesh used for discretizing the 2D simulation geometry. And the value assigned in the middle node of the triangle represents the permeability value for the whole triangular element. But for simplicity, in the remaining part of this section, we will imagine or assume the $2D$ geometry is discritized by a square grid with the value for each square element assigned in the middle of the element. This explanation allow one to easily visualize this permeability space because it is defined by the $x,y$ co-ordinates  system which most people are already familiar with. The model order reduction for the permeability field discussed here is concerned with reducing the volume of permeability data within the permeability space. Consider for instance a small square which is discretized to contain four squares such that $4$ data values are defined in the middle of the four squares. The permeability space of $4$ can be reduced to a smaller value of $1$ through an averaging process such as arithmetic, harmonic or geometric mean. This type of reduction is the focus of this paper, albeit the main focus of this paper is achieving this reduction through singular value decomposition in the principal component spaces.

@@

@@

The first section briefly introduced the rational behind upscaling. This is a subject which has interested the academic community for over $7$ decades.  \citet{Cardwell_1945} investigated arithmetic and weighted averages as upscaling techniques for calculating a single equivalent permeability for an oil reservoir with varying permeabilities. \citet{King1996} investigated the use of renormalization for upscaling. And \citet{Yeo2001} discussed the accuracy of renormalization method for upscaling 2 dimension hydraulic conductivity fields.  \citet{Renard_1997} provided a comprehensive review of the various techniques for upscaling in \citeyear{Renard_1997}. \citet{Christie_2001SPE10Model} provided $2$ sets of problems to compare the upscaling and upgridding techniques of different simulators.

>>>>

\newcommand{\JGnote}[1]{\fbox{\parbox{\textwidth}{ \color{red} JG Note $\Rightarrow$ #1}}}
\newcommand{\BLnote}[1]{\fbox{\parbox{\textwidth}{ \color{green} BL Note $\Rightarrow$ #1}}}
\newcommand{\red}{\textcolor{red}}
\newcommand{\blue}{\textcolor{blue}}
\newcommand{\green}{\textcolor{green}}
\newcommand{\frc}{\displaystyle\frac}
\newcommand{\PN}[2][error]{P$_{#1}$DG-P$_{#2}$}
\newcommand{\PNDG}[2][error]{P$_{#1}$DG-P$_{#2}$DG}
\newcommand{\eg}{{\it e.g., }} 
\newcommand{\ie}{{\it i.e., }}
\newcommand{\st}{{\it s.t., }}

@@

In Section 2, the basic mathematical explanation of SVD and its associated properties are presented. Then, the concept of spaces\red{(spaces ... could you be more specific?)} is explained from the engineer's perspective and also from the mathematician's perspective \red{(Please do rephrase this sentence ... engineers/mathematician perspective is not appropriate for a scientific journal ...)}. The objective of this explanation is to assist both parties in \green{visualizing} the relevant spaces and performing the necessary transformation within those spaces. Finally, the section \red{(which section?)} describes model order space reduction for the permeability field within the principal component space. Section 3, starts by describing the high resolution base case on which the four upscaling techniques are modelled. The the simulation run which is similar for all the cases/models is described. A brief summary of the pre-processing step for each of the models is also presented. Section 4 provides the results and some discussion on the results of the simulations. It als gives further intepretation to the result obtained from the SVD upscaling technique. Finally, section 5 provides the conclusion.

\section{Singular Value Decomposition, SVD, and its application for reducing the order of a permeability field}\label{section:svd}
\subsection{A Brief Introduction to SVD}\label{subsection:svd_brief}
Given a matrix $\mathbf{A}\in\mathcal{R}^{m\times n}$, the singular value decomposition (SVD) is a method for factorizing $\mathbf{A}$, such that,
\begin{equation}
  \mathbf{A} = \mathbf{U}\mathbf{\Sigma}\mathbf{V}^{\intercal}.
\end{equation}
where $\mathbf{U}=\left[\underline{u}_{1},\underline{u}_{2},\cdots,\underline{u}_{m}\right]\in\mathcal{R}^{m\times n}$ and $\mathbf{V}=\left[\underline{v}_{1},\underline{v}_{2},\cdots,\underline{v}_{n}\right]\in\mathcal{R}^{m\times n}$ are orthogonal but not necessarily the same. $\Sigma$ is a diagonal matrix, \ie $\Sigma=\text{diag}\left(\sigma_{1},\sigma_{2},\cdots,\sigma_{r}\right)\in\mathcal{R}^{m\times n}$ and $r=\min\left\{m,n\right\}$ is the rank of $\mathbf{A}$, with $\sigma_{1}\ge\sigma_{2}\ge\cdots\sigma_{r}\ge 0$. The diagonal entries of $\Sigma$ $\left(\ie \sigma_{i}, i=1,r\right)$ are called singular values of $\mathbf{A}$. If $\mathbf{A}$ is a square matrix then $\mathbf{U}$, $\Sigma$ and $\mathbf{V}$ are also square matrices. Also, as $\mathbf{U}$ and $\mathbf{V}$ are orthogonals and unitary, 
\begin{equation}
  \mathbf{U}^{\intercal}\mathbf{U} = \mathbf{V}^{\intercal}\mathbf{V} = \mathbf{I},
\end{equation}
where $\mathbf{I}$ is an identity matrix. However, if $\mathbf{A}$ is an $m\times n$ matrix, $\mathbf{U}$ and $\mathbf{V}$ still remain as square matrices, albeit with different matricial sizes. $\mathbf{U}$ is an $m\times m$ while $\mathbf{V}$ is an $n\times n$. Another relevant property of $\mathbf{U}$ and $\mathbf{V}$ since they are square (irrespective of whether $\mathbf{A}$ is square or rectangular) is that
\begin{equation}
  \mathbf{U}^{-1} = \mathbf{U}^{\intercal} \;\;\;\text{ and }\;\;\; \mathbf{V}^{-1} = \mathbf{V}^{\intercal},
\end{equation}
and if $\mathbf{A}$ is an $m \times n$ matrix thus $\Sigma$ is a diagonal $m \times n$ matrix. For a rectangular matrix, the singular values fill the first $r$ places on the main diagonal of $\Sigma$. A benefit of the SVD is the orthogonality of the columns in $\mathbf{U}$ and $\mathbf{V}$. The columns of $\mathbf{U}$ are known as the left singular vectors while the columns in $\mathbf{V}$ are known as the right singular vectors. A simple consideration of the benefit of this is the ease with which it allows the inverse of square matrix $\mathbf{A}$ to be calculated. This example is presented thus:
\begin{equation}
 \mathbf{A}x = b
\end{equation}
\begin{equation}
 \mathbf{U} \Sigma \mathbf{V}^{\intercal} x = b
\end{equation}
\begin{equation}
 x = \mathbf{V} \Sigma^{-1} \mathbf{U}^{\intercal} b
\end{equation}
\red{(I guess we have talked about this on Friday ... make it clear that matrix inversion is an (useful?) example of SVD ... also explain step-by-step this matricial equations above ... remember ... every single equation must be explained ...)}

\subsection{Principal Component Spaces and Model Order Reduction for the permeability field}\label{subsection:pcspaces}
\JGnote{As we talked on Friday: (a)make sure you explain what permeability is before starting to operating on it; (b) $P0DG$ could be easily replaced by saying that the quantity of interest is allocated in the centre of the element/cell -- this make it more general; (c) Just reading it through, we can summarise the {\it procedure} as (i) SVD of matrix of permeability scalar quantities (\ie upscaling) and (ii) linear interpolation of the upscaled/homogeneised quantities upon a new mesh grid resolution (\ie upgriding). Whereas the first is the object of the interest for this study, the second gives an overall motivation for the work ... thus, make sure this is clear within the text, \ie upscaling $\rightarrow$ upgriding.}

\subsubsection{Visualizing the permeability space}\label{subsubsection:visualization_permspace}
In 3D geometry, permeability is a tensor quantity, but for the purpose of the following explanation (and the remaining part of this paper), a 2D geometry is assumed where permeability is a scalar quantity. Furthermore, the simulator prescribes the permeability field using $P0DG$ element type, which means that permeability values are assigned in the middle node of the triangular unstructured mesh used for discretizing the 2D simulation geometry. And the value assigned in the middle node of the triangle represents the permeability value for the whole triangular element. But for simplicity, in the remaining part of this section, we will imagine or assume the $2D$ geometry is discritized by a square grid with the value for each square element assigned in the middle of the element. This explanation allow one to easily visualize this permeability space because it is defined by the $x,y$ co-ordinates  system which most people are already familiar with. The model order reduction for the permeability field discussed here is concerned with reducing the volume of permeability data within the permeability space. Consider for instance a small square which is discretized to contain four squares such that $4$ data values are defined in the middle of the four squares. The permeability space of $4$ can be reduced to a smaller value of $1$ through an averaging process such as arithmetic, harmonic or geometric mean. This type of reduction is the focus of this paper, albeit the main focus of this paper is achieving this reduction through singular value decomposition in the principal component spaces.

@@

\JGnote{Maybe you should consider describing the procedure above as pseudocode within a table.}

@@

The first section \red{(Are you referring to Section 1 -- Intro or Section 2?)} briefly introduced the rational behind upscaling. Upscaling methods replace discrete geological and fluid properties of detailed high resolution domains with coarse descriptions (low resolution) of these properties \cite{Vereecken_2007}. This is a subject which has interested the academic community for over $7$ decades.  \citet{Cardwell_1945} investigated arithmetic and weighted averages as upscaling techniques for calculating a single equivalent permeability for an oil reservoir with varying permeabilities. \citet{King1996} investigated the use of renormalization for upscaling. And \citet{Yeo2001} discussed the accuracy of renormalization method for upscaling 2D hydraulic conductivity fields.  \citet{Renard_1997} provided a comprehensive review of the various techniques for upscaling. \citet{Christie_2001SPE10Model} provided $2$ sets of problems to compare the upscaling and upgridding techniques of different simulators. \red{(And what are the main conclusions/contributions of these papers?)}
